### 2020-02-25 Writing/disparity_review.tex
Edits to part 2

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
\caption{
    The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation and can be linked to developmental constraints; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
    These categories are not independent and many disparity studies will cover more than one.
    %@@@ Add disparity and evo-devo
}
\label{Fig:disparity}
\end{figure}

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Foote1989-fd, Deline2018-le};
	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep{Raup1961-vx,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
However, if the question focuses on comparing wings between bats and birds, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
However, if the focus is on convergence between different bats and birds, it would be preferable to use traits that have facilitated flight in both groups (e.g. digit length, integumentary system, etc.).
Where there is any doubt about the appropriate traits to choose, it may be preferable to use different kinds of data for the same feature to determine whether they capture the same pattern of disparity.

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.

Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Brazeau2017-kg, Palci2018-ni}.
Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
Therefore, trait data should be collected with the question in mind, or the question asked should be tailored to the limits of the data available.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_data.pdf}
\caption{
    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence or absence data) or as continuous data.
    Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
    These measurements can then be mathematically transformed (logarithm transforms, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
    Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.
}
\label{Fig:data}
\end{figure}

@@

The concept of what the morphospace \textit{is} can differ between authors and studies but is generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points.
In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combination of traits.
We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e. the morphospace contains all the observed combinations of traits (see appendix \ref{section:appendixmorphospace}).
Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
If the morphospace is transformed mathematically (i.e. the dataset is transformed using logs, ordinations, distance matrices, etc...) it remains the morphospace as long as all the observation and traits remain in the datatset.

>>>>

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
\caption{
    The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
    These categories are not independent and many disparity studies will cover more than one.
    %@@@ Add disparity and evo-devo
    % NC: I removed the mention of constraints because I hate the idea it's all constraints - there are other things at play why are we only mentioning constraints? (I suspect Anjali or Vera put that in?) Also it didn't fit with the other examples which don't mention mechanisms...
}
\label{Fig:disparity}
\end{figure}

@@

\noindent Disparity analyses are based on traits, but traits can be characterised in a number of ways:
	1) discrete morphological characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g.][]{Foote1989-fd, Deline2018-le};
	2) continuous measurements of features (e.g. lengths) \citep[e.g][]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[e.g.][]{Cooney2017-ly}, Fourier coefficients \citep[e.g.][]{Foote1995-do, Spriggs2018-nu} or model-based descriptors \citep[e.g.][]{Raup1961-vx,Saunders2004} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked using those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

For example, if investigating variation of bat wing shapes, both homologous landmarks and continuous measurements of bones may be appropriate to capture patterns of wing variation.
If the question focuses on comparing wings between bats and birds, however, different measurements might be more appropriate depending on the specific question: if the focus is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
However, if the focus is on convergence between different bats and birds, it would be preferable to use traits that have facilitated flight in both groups (e.g. digit length, integumentary system, etc.).
Where there is any doubt about the appropriate traits to choose, it may be preferable to use several different kinds of data for the same feature to determine whether they capture the same pattern of disparity.

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character are commonly recycled from phylogenetic studies \citep[e.g.][]{Brusatte2008-fx,Close2015-qi,Wright2017-jo}.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017-cf}.
This effect of the anatomical feature used on disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018-le}.

Trait data suffer from the same shortcomings as most biological datasets -- data can be missing, non-overlapping, hierarchical, inapplicable, ambiguous, polymorphic, correlated, or there may be an insufficient sample size \citep{Brazeau2017-kg, Palci2018-ni}.
Biological phenomena such as allometry and sexual dimorphism may also influence trait data.
More mundanely, data collection is constrained by the time and money available, making collating a ``perfect'' dataset difficult.
Ultimately, disparity analyses characterise the data, and subsamples of the universe of possible data may not have the power to uncover holistic patterns of disparity.
Therefore, trait data should be collected with the scientific question in mind, or the question asked should be tailored to the limits of the data available.

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_data.pdf}
\caption{
    Major routes to obtain morphological data for disparity analyses. Data can be collected as discrete trait observations (e.g. presence-absence data) or as continuous data.
    Continuous data can be collected by various methods including linear measurements and landmark coordinates or contours (curves).
    These measurements can then be mathematically transformed (logarithmic transformations, scaling, Procrustes superimposition, elliptic Fourier transforms, etc.).
    Regardless of the method, data collection produces a trait matrix where the observed traits constitute columns and the studied elements (generally taxa or OTUs) the rows.
}
\label{Fig:data}
\end{figure}

@@

The concept of what the morphospace \textit{is} can differ among authors and studies but is generally used in the common (i.e. non-mathematical) sense of a space containing all possible morphologies.
For the purpose of this review, we use a generalised version of the definition of \cite{mitteroecker2009concept} where the morphospace is a \textit{p}-dimensional space (or Q-space) containing \textit{n} points.
In other words, mathematically, the morphospace is the mathematical object that contains the trait values (\textit{p}, the axes of the space) for all the observations (\textit{n}, the specimens); or, biologically, the morphospace contains all the observed combinations of traits.
We generalise this definition to any linear transformation ($V \to W$) of this space that retains the biological definition, i.e. the morphospace contains all the observed combinations of traits (see appendix \ref{section:appendixmorphospace}).
Note that here the morphospace does not need to be Euclidean (although this is often desired or assumed - see section \ref{section:ordination} below).
If the morphospace is transformed mathematically (i.e. the dataset is transformed using logarithms, ordinations, distance matrices, etc.) it remains the morphospace provided all the observation and traits remain in the datatset.
